---
title: "Assignment 5: Data Visualization"
author: "Natasha Jacob"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE, echo = T)
```

```{r}
#1 Checking working directory
getwd()
# Loading tidyverse and cowplot packages 
library(tidyverse)
library(cowplot)
# Uploading the processed data files 
PeterPaul_Processed <- read.csv("../Environmental_Data_Analytics_2022/Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",
                                stringsAsFactors = TRUE)
Litter_Processed <- read.csv("../Environmental_Data_Analytics_2022/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", 
                             stringsAsFactors = TRUE)

#2 Changing date to date format 
PeterPaul_Processed$sampledate <- as.Date(PeterPaul_Processed$sampledate, format = "%Y-%m-%d")
Litter_Processed$collectDate <- as.Date(Litter_Processed$collectDate, format = "%Y-%m-%d")

#converting month as a factor 
PeterPaul_Processed$month <- as.factor(PeterPaul_Processed$month)

#Checking the class of month
class(PeterPaul_Processed$month)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3 Building a theme 
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4 Plotting total phosphorous by phosphate for Peter and Paul Lakes
Total_phosphorous_by_phosphate <- ggplot(PeterPaul_Processed, 
                                         aes(x = po4, y = tp_ug)) +
  geom_point(aes(color = lakename)) +
  geom_smooth(method = lm, color = "black") +
  xlim(0,50) + ylim(0, 150) +
   ylab(expression("Phosphate concentration")) +
  xlab(expression("Total phosphorous concentration")) +
  mytheme
print(Total_phosphorous_by_phosphate)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5 Creating three separate boxplots of month versus temperature, total phosphorous and total nitrogen. Legends were not generated for these plots. 

  Temperature_boxplot <- ggplot(PeterPaul_Processed, 
                                aes(x = month, y = temperature_C)) +
    geom_boxplot(aes(color = lakename)) + 
   scale_x_discrete(limits = factor("5":"11")) +
  theme_classic(base_size = 10) + 
  ylab(expression("Temperature (in Celcius)")) +
  theme(axis.text = element_text(color = "black"), legend.position = "none") 
print(Temperature_boxplot)
  
  Phosphorous_boxplot <- ggplot(PeterPaul_Processed, 
                                aes(x = month, y = tp_ug)) +
    geom_boxplot(aes(color = lakename)) + 
     scale_x_discrete(limits = factor("5":"11")) +
    theme_classic(base_size = 10) + 
    ylab(expression("Total Phosphorous (ug)")) +
  theme(axis.text = element_text(color = "black"), legend.position = "none")
  print(Phosphorous_boxplot)
  
  Nitrogen_boxplot <- ggplot(PeterPaul_Processed, aes(x = month, y = tn_ug)) +
    geom_boxplot(aes(color = lakename))  +
     scale_x_discrete(limits = factor("5":"11")) + 
    ylab(expression("Total Nitrogen (ug)")) +
    theme_classic(base_size = 10) + 
  theme(axis.text = element_text(color = "black"), legend.position = "none")
  print(Nitrogen_boxplot)
  
#Creating a separate boxplot of month vs temperature to obtain the legend. (Notice that legend position is specified as "top" in the code)   
  Temperature_boxplot1 <- ggplot(PeterPaul_Processed, 
                                 aes(x = month, y = temperature_C)) +
    geom_boxplot(aes(color = lakename)) + 
   scale_x_discrete(limits = factor("5":"11")) +
  theme_classic(base_size = 10) +
  theme(axis.text = element_text(color = "black"), legend.position = "bottom") 
  
#Using the get_legend() function to obtain the legend for the final combined graph 
NTL_legend <- get_legend(Temperature_boxplot1)

#Combining the three boxplots (without the legend)
plot1 <- plot_grid(Temperature_boxplot, Phosphorous_boxplot, Nitrogen_boxplot, labels = c("Temperature", "Phosphorous", "Nitrogen") , ncol = 3, rel_heights = c(3, 0.3))

#Combining three boxplots with one legend present 
Combined_plot <- plot_grid(plot1, NTL_legend, ncol = 1, nrow = 2, rel_heights = c(5, 0.3))
Combined_plot

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature plot: A pattern of normally distributed values can be observed in the temperature plot (for both lakes) as the temperature seems to increase starting from May, reaching its peak in July and a slow decline starting from August can be observed. The lowest recorded temperture, however, can be observed in the month of November. 
Phosphorous plot: A steady increase in phosphorous content can be observed for Peter Lake. However, a steady decrease in phosphorous content up until August can be seen in Paul Lake after which the phosphorous content seems to increase in September. 
Nitrogen plot: There isn't much variation in Nirogen levels observed in Peter and Paul Lakes excpet for the minor incline for Peter Lake observed in the month of August. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6 Plotting dry mass of needle by Date, separated by NLCD class (using a color aesthetic)
DryMass_by_date_nlcd <- ggplot(subset(Litter_Processed, functionalGroup == "Needles"),
       aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  mytheme
print(DryMass_by_date_nlcd)

#7 Plotting dry mass of needle by Date with NLCD classes separated into three facets
DryMass_by_date_facet <- ggplot(subset(Litter_Processed, functionalGroup == "Needles"),
       aes(x = collectDate, y = dryMass)) +
  facet_wrap(vars(nlcdClass), nrow = 3) +
  geom_point() +
  mytheme
print(DryMass_by_date_facet)
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 where we plotted dry mass of needle by Date with NLCD classes separated into three facets is a more effective plot to visualise data since the facets make it easier to observe the seperate nlcd classes. In plot #6, it is difficult to visualise data since the points appear clustered. 
